### 2020-08-05 ART_Czarnik/Czarnik-PU.tex
P. Polak correctons added

\begin{artengenv}{Szymon Czarnik}
	{How Much Truth Is in Stereotypes?}
	{How Much Truth Is in Stereotypes?}
	{How Much Truth Is in Stereotypes?}
	{Institute of Sociology, Jagiellonian Universitys}
	{Stereotype accuracy is a contentious topic. Part of the problem is that typically stereotypes are generic statements whose truth status is unclear due to the fact that they are ill-defined quantitatively. The article focuses on the epistemic aspect of stereotypical beliefs. In the ongoing debate, I side with those who argue against stereotypes being wrong or inaccurate by virtue of definition alone. I propose that, when possible, stereotype accuracy should be assessed in probabilistic terms by inspecting how likely a generic statement is to be true when applied to individual(s) representative of the relevant group(s). This approach applies equally well to investigating the actual and the perceived accuracy of stereotypes.}
	{stereotypes, stereotype accuracy, epistemology, probabilistic approach.}

\vspace{2\baselineskip}
In this paper, I propose that stereotypes are best understood as generic statements whose truth status is unclear. They are expressed by asserting that ``members of group A have trait T,'' or ``Members of group A are higher/lower on trait T than members of group B.'' Such utterances lack proper quantification and invite alternative interpretations. One such interpretation is that stereotypes speak of all group members without exception which leads to the idea that stereotypes should be considered false by definition, which in turn leads to the opposition to stereotype accuracy research as, at best, a misguided endeavor. This opposition, however, is unfounded on epistemic grounds once we explicitly recognize generic nature of stereotypes. Thus after outlining the long-lasting debate, I proceed to discuss the problems inherent in measuring the accuracy of stereotypes, and then advocate a probabilistic approach, i.e. assessment of probabilities that individuals conform to the generic rule posited by a stereotype. As a practical illustration, I show a number of such probabilistic assessments based on data from representative samples. Finally, I propose that the very same probabilistic approach should be used to directly investigate how accurate people perceive their own stereotypes to be. This in turn would provide an easy and straightforward method of comparing perceived and actual accuracy of stereotypes, acknowledging a possibility that stereotypical perceptions may both over- and underestimate actual differences.

@@

\enlargethispage{.7\baselineskip}
In daily life conversations, calling somebody's remark a stereotype is surely a form of criticism. In common parlance, as well as in vast sociological and psychological literature, the term ‘stereotype' implies untruth, or inaccuracy of the assertion. A general statement like ``Scots are cheap'' may be picked apart as untrue in the sense that on average Scots are no more stingy than members of any other ethnic group. Or, even if the statement had some kernel of truth in it, one could be brought to task for inaccurately applying it to pass a judgment on a particular individual who may happen not to conform to this general rule.\footnote{Far from being limited to stereotypical beliefs, the problem of group-to-individual generalizability may pose a serious threat to scientific research involving human subjects
%\label{ref:RNDTlfuAJ4ACM}(Fisher, Medaglia and Jeronimus, 2018).
\parencite[][]{fisher_lack_2018}.%
} Apart from being misguided epistemologically, the utterance may also be criticized on moral grounds as an attempt at a wholesale vilification of a certain group of people. Thus the presumed wrongness of stereotypes may consist in epistemic and/or moral failure 
%\label{ref:RNDo2Xq32gcjA}(Beeghly, 2014).
\parencite[][]{beeghly_seeing_2014}. %
 It is solely the former aspect of stereotypes that we shall be concerned with here.\footnote{Some authors reduce stereotypes to the epistemic aspect alone. Thus Banaji claims that ``[f]rom the earliest use of the term in psychology, stereotypes have been regarded as the cognitive (thought) as opposed to the a[FB00?]ective (feeling) component of mental representations of social groups. As such, the construct is tied to but di[FB00?]erentiated from the concept of attitude, preference, or liking as well as the concept of discrimination'' 
%\label{ref:RNDHS9MlJPRnC}(Banaji, 2001, p.15101).
\parencite[][p.15101]{banaji_stereotypes_2001}. %
 }

\section{Resistance to research on stereotype accuracy}
Before we proceed further it should be acknowledged that, for a good reason, the ethical aspect looms large in the scholarly literature on the topic. In psychology, stereotypes are studied as cognitive components of prejudice, which may be brought about by and contribute to inter-group hostility
%\label{ref:RND4lT4IExI0w}(Bar-Tal, 1989);
\parencite[][]{bar-tal_delegitimization_1989}; %
 in sociology, they are recognized as a device that may adversely affect targeted groups by means of self-fulfilling prophecy 
%\label{ref:RNDq8zLOYvax2}(Merton, 1948);
\parencite[][]{merton_self-fulfilling_1948}; %
 in criminology, they are discussed in the context of racial/ethnic profiling in police operations 
%\label{ref:RNDFHF118jk89}(Schauer, 2003);
\parencite[][]{schauer_profiles_2003}; %
 in historical studies, they are investigated as part and parcel of totalitarian propaganda 
%\label{ref:RND3zPpNfYi3y}(Werth, 2011; USHMM (United States Holocaust Memorial Museum),
\parencites[][]{werth_dekulakisation_2011}[][]{ushmm_united_states_holocaust_memorial_museum_defining_nodate}, %
 2020); in economics, they are incorporated in models of statistical discrimination in the labor market 
%\label{ref:RNDgoefvun65q}(Arrow, 1971; Phelps, 1972).
\parencites[][]{arrow_theory_1971}[][]{phelps_statistical_1972}. %
 It goes without saying that after Adorno had linked stereotypic thinking with fascist belief systems the concept became even more anathema 
%\label{ref:RNDHdzxZxrjhG}(Jones and Colman, 2001).
\parencite[][]{jones_stereotypes_2001}.%

@@

\enlargethispage{.5\baselineskip}
\section[Problems in measuring the accuracy of stereotypes]{Problems in measuring the accuracy of stereotypes\footnote{It should be stressed that what I am interested in here is the extent to which stereotypical judgments as such are correct. It is a related but separate issue whether the process by which people arrive at their judgments is correct. For example, someone may arrive at a correct estimate of honesty in a given group of people, even though he overestimates honesty of people in general while underestimating honesty of that particular group relative to general public. See
%\label{ref:RNDq1DptTqYi9}(Jussim et al., 2016)
\parencite[][]{jussim_stereotype_2016} %
 for a discussion of processual models of judgment by Cronbach 
%\label{ref:RNDqEAFrZBRPd}(1955)
\parencite*[][]{cronbach_processes_1955} %
 and Kenny 
%\label{ref:RNDlQ5sgtU4cA}(Kenny, 1994).
\parencite[][]{kenny_interpersonal_1994}. %
 }}
Assessing stereotype (in)accuracy involves three obvious steps
%\label{ref:RNDr97wsTQtKY}(Jussim, Crawford and Rubinstein, 2015):
\parencite[][]{jussim_stereotype_2015}:%

@@

Even if many group differences turn out to be real rather than imaginary, bigots—to reiterate Kenny
%\label{ref:RND84Zh1FAM8u}(Kenny, 1994, p.212)
\parencite[][p.212]{kenny_interpersonal_1994}%
—``should not take any comfort in these conclusions.''\footnote{Bigots should also be well advised to recognize that they themselves are a strongly stereotyped group, and these stereotypes may turn out to be not so far from the truth, either. } Such differences leave space—in some instances a whole lot of space—for exceptions. They also evolve: they may grow bigger, but equally likely they may grow smaller, vanish or even be reversed. In any case, to minimize the stereotype-based discrimination, it is always advisable to advocate the individualistic approach, i.e. to explain why the best decisions are the ones based—whenever possible and to whatever extent feasible—on person's individual characteristics rather than on the average characteristics of the groups individuals belong to.

>>>>

\begin{artengenv}{Szymon Czarnik}
	{How much truth is in stereotypes?}
	{How much truth is in stereotypes?}
	{How much truth is in stereotypes?}
	{Institute of Sociology, Jagiellonian Universitys}
	{Stereotype accuracy is a contentious topic. Part of the problem is that typically stereotypes are generic statements whose truth status is unclear due to the fact that they are ill-defined quantitatively. The article focuses on the epistemic aspect of stereotypical beliefs. In the ongoing debate, I side with those who argue against stereotypes being wrong or inaccurate by virtue of definition alone. I propose that, when possible, stereotype accuracy should be assessed in probabilistic terms by inspecting how likely a generic statement is to be true when applied to individual(s) representative of the relevant group(s). This approach applies equally well to investigating the actual and the perceived accuracy of stereotypes.}
	{stereotypes, stereotype accuracy, epistemology, probabilistic approach.}

\vspace{2\baselineskip}
\lettrine[loversize=0.13,lines=2,lraise=-0.05,nindent=0em,findent=0.2pt]%
{I}{}n this paper, I propose that stereotypes are best understood as generic statements whose truth status is unclear. They are expressed by asserting that ``members of group A have trait T,'' or ``Members of group A are higher/lower on trait T than members of group B.'' Such utterances lack proper quantification and invite alternative interpretations. One such interpretation is that stereotypes speak of all group members without exception which leads to the idea that stereotypes should be considered false by definition, which in turn leads to the opposition to stereotype accuracy research as, at best, a misguided endeavor. This opposition, however, is unfounded on epistemic grounds once we explicitly recognize generic nature of stereotypes. Thus after outlining the long-lasting debate, I proceed to discuss the problems inherent in measuring the accuracy of stereotypes, and then advocate a probabilistic approach, i.e. assessment of probabilities that individuals conform to the generic rule posited by a stereotype. As a practical illustration, I show a number of such probabilistic assessments based on data from representative samples. Finally, I propose that the very same probabilistic approach should be used to directly investigate how accurate people perceive their own stereotypes to be. This in turn would provide an easy and straightforward method of comparing perceived and actual accuracy of stereotypes, acknowledging a possibility that stereotypical perceptions may both over- and underestimate actual differences.

@@

\enlargethispage{.7\baselineskip}
In daily life conversations, calling somebody's remark a stereotype is surely a form of criticism. In common parlance, as well as in vast sociological and psychological literature, the term ‘stereotype' implies untruth, or inaccuracy of the assertion. A general statement like ``Scots are cheap'' may be picked apart as untrue in the sense that on average Scots are no more stingy than members of any other ethnic group. Or, even if the statement had some kernel of truth in it, one could be brought to task for inaccurately applying it to pass a judgment on a particular individual who may happen not to conform to this general rule.\footnote{Far from being limited to stereotypical beliefs, the problem of group-to-individual generalizability may pose a serious threat to scientific research involving human subjects
%\label{ref:RNDTlfuAJ4ACM}(Fisher, Medaglia and Jeronimus, 2018).
\parencite[][]{fisher_lack_2018}.%
} Apart from being misguided epistemologically, the utterance may also be criticized on moral grounds as an attempt at a wholesale vilification of a certain group of people. Thus the presumed wrongness of stereotypes may consist in epistemic and/or moral failure 
%\label{ref:RNDo2Xq32gcjA}(Beeghly, 2014).
\parencite[][]{beeghly_seeing_2014}. %
 It is solely the former aspect of stereotypes that we shall be concerned with here.\footnote{Some authors reduce stereotypes to the epistemic aspect alone. Thus Banaji claims that ``[f]rom the earliest use of the term in psychology, stereotypes have been regarded as the cognitive (thought) as opposed to the affective (feeling) component of mental representations of social groups. As such, the construct is tied to but differentiated from the concept of attitude, preference, or liking as well as the concept of discrimination'' 
%\label{ref:RNDHS9MlJPRnC}(Banaji, 2001, p.15101).
\parencite[][p.15101]{banaji_stereotypes_2001}. %
 }

\section{Resistance to research on stereotype accuracy}
Before we proceed further it should be acknowledged that, for a good reason, the ethical aspect looms large in the scholarly literature on the topic. In psychology, stereotypes are studied as cognitive components of prejudice, which may be brought about by and contribute to inter-group hostility
%\label{ref:RND4lT4IExI0w}(Bar-Tal, 1989);
\parencite[][]{bar-tal_delegitimization_1989}; %
 in sociology, they are recognized as a device that may adversely affect targeted groups by means of self-fulfilling prophecy 
%\label{ref:RNDq8zLOYvax2}(Merton, 1948);
\parencite[][]{merton_self-fulfilling_1948}; %
 in criminology, they are discussed in the context of racial/ethnic profiling in police operations 
%\label{ref:RNDFHF118jk89}(Schauer, 2003);
\parencite[][]{schauer_profiles_2003}; %
 in historical studies, they are investigated as part and parcel of totalitarian propaganda 
%\label{ref:RND3zPpNfYi3y}(Werth, 2011; USHMM (United States Holocaust Memorial Museum),
\parencites[][]{werth_dekulakisation_2011}[][]{ushmm_united_states_holocaust_memorial_museum_defining_nodate}%
; in economics, they are incorporated in models of statistical discrimination in the labor market 
%\label{ref:RNDgoefvun65q}(Arrow, 1971; Phelps, 1972).
\parencites[][]{arrow_theory_1971}[][]{phelps_statistical_1972}. %
 It goes without saying that after Adorno had linked stereotypic thinking with fascist belief systems the concept became even more anathema 
%\label{ref:RNDHdzxZxrjhG}(Jones and Colman, 2001).
\parencite[][]{jones_stereotypes_2001}.%

@@

\enlargethispage{.5\baselineskip}
\section[Problems in measuring the accuracy of stereotypes]{Problems in measuring the accuracy of stereotypes\footnote{It should be stressed that what I am interested in here is the extent to which stereotypical judgments as such are correct. It is a related but separate issue whether the process by which people arrive at their judgments is correct. For example, someone may arrive at a correct estimate of honesty in a given group of people, even though he overestimates honesty of people in general while underestimating honesty of that particular group relative to general public. See
%\label{ref:RNDq1DptTqYi9}(Jussim et al., 2016)
\parencite[][]{jussim_stereotype_2016} %
 for a discussion of processual models of judgment by Cronbach 
%\label{ref:RNDqEAFrZBRPd}(1955)
\parencite*[][]{cronbach_processes_1955} %
 and Kenny 
%\label{ref:RNDlQ5sgtU4cA}(Kenny, 1994).
\parencite*[][]{kenny_interpersonal_1994}. %
 }}
Assessing stereotype (in)accuracy involves three obvious steps
%\label{ref:RNDr97wsTQtKY}(Jussim, Crawford and Rubinstein, 2015):
\parencite[][]{jussim_stereotype_2015}:%

@@

Even if many group differences turn out to be real rather than imaginary, bigots—to reiterate Kenny
%\label{ref:RND84Zh1FAM8u}(Kenny, 1994, p.212)
\parencite*[][p.212]{kenny_interpersonal_1994}%
—``should not take any comfort in these conclusions.''\footnote{Bigots should also be well advised to recognize that they themselves are a strongly stereotyped group, and these stereotypes may turn out to be not so far from the truth, either. } Such differences leave space—in some instances a whole lot of space—for exceptions. They also evolve: they may grow bigger, but equally likely they may grow smaller, vanish or even be reversed. In any case, to minimize the stereotype-based discrimination, it is always advisable to advocate the individualistic approach, i.e. to explain why the best decisions are the ones based—whenever possible and to whatever extent feasible—on person's individual characteristics rather than on the average characteristics of the groups individuals belong to.
